---
title: "Data 607 Project 1"
author: "Steven Ellingson"
date: "September 10, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Pull in data.  Instead of pulling in line by line, I'm going to just pull it into one big chunk.
```{r, message=FALSE, warning=FALSE}
library(stringr)
library(dplyr)
raw.data <- readChar('tournamentinfo.txt',file.info('tournamentinfo.txt')$size)

```
Get rid of the ----'s and pull each player's data onto one line
```{r}
data.raw.2 <-unlist(str_replace_all(str_replace_all(raw.data,'\n',''),'-{89}','\n'))
data.lines <- unlist(strsplit(data.raw.2,'\n'))
head(data.lines)
```

Remove the top line as it's blank.  Split out some fields that are deliminated in different ways and clean up results
```{r}
data.lines <- data.lines[-1]
data.lines <- str_replace_all(data.lines,'\r','')
data.lines.2 <- str_replace_all(data.lines,'/','|')
data.lines.2 <- str_replace_all(data.lines.2,'->','|')
data.lines.2 <- str_trim(str_replace_all(data.lines.2,' *\\| *','|'))
head(data.lines.2)
```

Now each of the lines have the fields separated by a |.  The headers don't all make sense but I'll leave them for now.
Let's pull just the data we need into a dataset.  I'll keep opponents as a comma separated list for now and pull the opponent's rating after.
```{r}
#This function will split out just the data we need and return it as a vector
pullData = function(v) {
  data = unlist(str_split(v, '\\|'))
  id = data[1]
  name = data[2]
  state = data[11]
  points = data[3]
  prerating = unlist(str_split(unlist(str_split(data[13], ' +'))[2], 'P'))[1]
  opponents = sapply(data[4:10],function(x) unlist(str_extract_all(x, '\\d+')))
  opponentString = str_replace_all(paste(opponents,collapse=","),'character\\(0\\)','')
  return(c(id,name,state,points,prerating,opponentString))
}

#Apply the function over our list to create a dataframe
data.lines.3 <- data.frame(do.call(rbind,lapply(data.lines.2, pullData)),stringsAsFactors = FALSE)

#Fix column names and remove the first line, as that is the header
colnames(data.lines.3) <- c('ID','Name','State','Points','Pre-Rating','OpponentList')
data.lines.3 = data.lines.3[-1, ]

#Now that we removed the header we can fix our numeric data.
data.lines.3$ID = type.convert(data.lines.3$ID)
data.lines.3['Pre-Rating'] = type.convert(data.lines.3[,'Pre-Rating'])
#Set the rows namesto the ID
rownames(data.lines.3) = data.lines.3$ID

head(data.lines.3)
```


OK, so now I just need a function to loop through the "OpponentList" and take an average of the opponent's prerating

```{r}
#This function will pull in a comma separated list and average the ratings for that list of player IDs
opponentAverage = function(l) {
  v = unlist(str_split(l,','))
  v = v[v != ""]
  v = type.convert(v)
  ratings = round(as.double(lapply(v, function(x) data.lines.3[x,'Pre-Rating'])), 1)
  return(mean(ratings))
}

#Apply over our data frame
data.lines.3$OpponentAverage <- as.double(lapply(data.lines.3$OpponentList,opponentAverage))

#Remove the ID and OpponenetList as we don't want those in our final data set.
data.final = select(data.lines.3,-c(ID,OpponentList))
head(data.final)
```

Looks pretty good! Let's save our csv.
```{r}
write.csv(data.final, 'chess_player_list.csv', row.names=FALSE)
```