---
title: "Data-607-Assignment-2"
author: "Steven Ellingson"
date: "September 5, 2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

I Listed some classic Nicolas Cage movies and had may siblings rate them on Google Sheets. If I had to do it over I would have picked some Marvel movies or something since my data ended up pretty sparse.  


```{r}
library(googlesheets)
library(RMySQL)
library(reshape2)
```


Connect to local DB.
redacted code will look like this:  
con = dbConnect(dbDriver("MySQL"), user='', password='', dbname='cuny', host='',port = 3306)
```{r, echo = FALSE}
#on.exit(dbDisconnect(con))
```
This is probably not best practice but I have all of the SQL inside of the R code. It was a bit of a challenge to get it all to work but it's nice that it's all together in one document, and it gets completely reset every time I change and re-run.

Drop Tables if they exist

```{r}
res <- dbSendQuery(con,'drop table if exists movie_critics')
res <- dbSendQuery(con,'drop table if exists movies')
res <- dbSendQuery(con,'drop table if exists movie_ratings')
```

Create table to list critics
```{r}
res <- dbSendQuery(con,"
create table movie_critics (
id int primary key auto_increment,
name varchar(100)
)")
```
Create table to list movies
```{r}
res <- dbSendQuery(con,"
create table movies (
id int primary key auto_increment,
name varchar(100)
)")
```
Create table of ratings
I created "rating" as a decimal in case someone put a half star.
```{r}
res <- dbSendQuery(con,"
create table movie_ratings(
id int primary key auto_increment,
critic_id int,
movie_id int,
rating decimal(14,4)
)")
```
Use stored token to get access to Google Sheets
```{r}
gs_auth(token = "googlesheets_token.rds")
```
Pull in the spreadsheet
```{r}
sheet <- gs_url('https://docs.google.com/spreadsheets/d/1eMTKJkwgP0mZup-25iF0u67EzgFUBEJYx3V3A5Khi_E')
sheet
```
List out the worksheets 
```{r}
gs_ws_ls(sheet)
```
First Insert the movie critics into the table
```{r}
query <- paste("INSERT INTO movie_critics (name) VALUES", paste("('",gs_ws_ls(sheet), "')", collapse = ", ", sep = ""), sep = "")
res <- dbSendQuery(con,query)
```
Now let's define a couple of functions to read our data and push into DB

This function should be run for each worksheet  
It will pull id for the critic, then use the "insert_rating" function to insert the rows
```{r}
read_worksheet <- function(ss, ws) {
  
  # Get ID from table for this critic
  res <- dbSendQuery(con, paste0("select id from movie_critics where lower(name) = '",ws,"'"))
  data <- fetch(res,n=-1)
  
  if (nrow(data) > 1) {
    print(paste("WARNING, MULITPLE CRITICS WITH NAME",ws))
  } else if (nrow(data) == 0) {
    print(paste("WARNING, NO CRITICS WITH NAME",ws))
  } else {
    critic_id = data$id[1]
  }
  
  ratings <- gs_read(ss, ws)
  apply(ratings, 1, function(x) insert_rating(ss,critic_id,x))
}
```
This function should be run for every each individual rating  
It will get the ids for the critic and movie and insert into DB
```{r}

insert_rating <- function(ss, critic_id, rating_row) {
  #Extract Movie and Rating from Row
  movie = rating_row['Movie']
  rating = rating_row['Rating']
  
  #Get ID for movie from table
  res <- dbSendQuery(con, paste0("select id from movies where lower(name) = '",tolower(movie),"'"))
  data <- fetch(res,n=-1)
  
  if (nrow(data) > 1) {
    print(paste("WARNING, MULITPLE MOVIES WITH NAME",movie))
  } else if (nrow(data) == 0) {
    #there is no entry for this movie, so add it!
    res <- dbSendQuery(con, paste0("insert into movies (name) values ('",movie,"')"))
  }
  
  # Run Query again in case it wasn't there
  res <- dbSendQuery(con, paste0("select id from movies where lower(name) = '",movie,"'"))
  data <- fetch(res, n=-1)
  # Assume the first one is the one we want
  movie_id = data$id[1]
  
  # I didn't give clear instructions on what to do if you haven't seen the movie
  # we'll assume anything that isn't a number should go in as null, but also give a warning
  
  if (is.na(as.numeric(rating))) {
    if (is.na(rating)) {
      print(paste0("Warning - N/A for ",critic_id,",",movie_id,': this is probably OK because it was N/A in google sheets'))
    } else {
      print(paste0("Warning - N/A for ",critic_id,",",movie_id,': "',rating,'" was changed to N/A!'))
    }
    # Either way, lets change to "null" so it can be inserted into DB
    rating = 'null'
  }
  

  res <- dbSendQuery(con, paste0("insert into movie_ratings (critic_id,movie_id,rating) values (",critic_id,",",movie_id,",",rating,")"))
  
}
```
Use the above functions to insert the data
```{r}
sapply(gs_ws_ls(sheet), function(x) read_worksheet(sheet,x))
```
OK, the warnings look reasonable (changing "Didn't Watch" and "N/A" to N/A seems ok) so lets pull our data into a data frame
```{r}
query = "select 
	mc.name as critic,
    m.name as movie,
    mr.rating
 from movie_ratings mr
 left join movies m on mr.movie_id = m.id
 left join movie_critics mc on mr.critic_id = mc.id"

res <- dbSendQuery(con, query)
data <- fetch(res, n=-1)
```
```{r}
str(data)
```
looks ok at a glance  Lets change it so the critics and movies are the row and column names
```{r}
data.pivot = dcast(data, movie ~ critic,value.var = "rating")
data.pivot
```
What the heck, there is a movie titled "Bangkok Dangerous2".  Since I don't believe that picture got a sequel I think something is wrong.
```{r}
data[data$movie=="Bangkok Dangerous2",]
```
I'm going to assume that Katie tried to put in a rating of 2 for "Bangkok Dangerous"
In real life I would fix this either in the spreadsheet itself or in the database, but for this exercize I'll just fix the data frame
```{r}
data[data$movie=="Bangkok Dangerous2",'rating'] = 2
data[data$movie=="Bangkok Dangerous2",'movie'] = "Bangkok Dangerous"
```
Let's try that again
```{r}
data.pivot = dcast(data, movie ~ critic,value.var = "rating")
data.pivot
```
lets change our row names to the movie titles
```{r}
row.names(data.pivot) = data.pivot$movie
data.pivot['movie']=NULL
```
Since we have our data we may as well take a look at some summary statistics

Check how many people watched each movie
```{r}
5-rowSums(is.na(data.pivot))
```
National Treasure and Face/Off were seen by everyone.  Only a couple saw Raising Arizona and The Rock, which is disappointing

How many responses did I get per person
```{r}
5-colSums(is.na(data.pivot))
```
I saw all of them, which makes sense since I picked the movies.  My younger siblings David and Katie are missing out.

Let's look at the average rating by movie
```{r}
round(rowMeans(data.pivot,na.rm=TRUE), 1)
```
More evidence that Katie and David should watch Raising Arizona and The Rock

And average rating by critic
```{r}
round(colMeans(data.pivot,na.rm=TRUE), 1)
```

Katie seems to be a pretty harsh critic. I wonder if we adjust the movie ratings based on the users average rating if that would adjust our results  
This is a pretty silly thing to do with such a sparse dataset, but it's my own fault I guess
```{r}
data.normalized <- round(sweep(data.pivot,2,colMeans(data.pivot,na.rm=TRUE),'/'), 1)
```
Let's now multiply by overall rating average to get it back on the same scale
```{r}
data.normalized <- data.normalized*mean(data$rating, na.rm=TRUE)
rbind(round(rowMeans(data.pivot,na.rm=TRUE), 1),
round(rowMeans(data.normalized,na.rm=TRUE), 1))
```
Looking at iot this way, action classics "The Rock and "Con Air" get taken down a notch while Bangkok Dangerous looks less pitiful

