---
title: "Data-607-Week-9"
author: "Steven Ellingson"
date: "10/27/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this assignment, I decided to take all of my personal ratings from rottentomatoes.com, and pull any New York Times reviews for those ratings.  

First, I will be pulling my personal ratings from rotten tomatoes.  Because it has an infinite scrolling page, I will use RSelenium to scroll to the the bottom of the page.

```{r}
library(RSelenium)
library(rvest)
library(httr)
library(jsonlite)
library(dplyr)
library(ggplot2)
```

```{r, eval=FALSE}
#This is the URL to my ratings. Rotten Tomatoes ratings are public, so I don't need to worry about loging in.
url <- "https://www.rottentomatoes.com/user/id/905861815/ratings"

#Open up Chrome Driver.  Because I'm not using a Docker container, YMMV here.
driver<- rsDriver()
remDr <- driver[["client"]]
remDr$navigate(url)

#scroll down 100 times, waiting for the page to load at each time
#In the future I should make this smarter - scroll to end instead of just a certain number of times.
for(i in 1:100){      
  remDr$executeScript(paste("scroll(0,",i*10000,");"))
  Sys.sleep(3)    
}
```

I'm going to pull just the titles and the ratings. In order to calculate the ratings, I count the number of "filled" stars, and the number of "half" stars.
````{r, eval=FALSE}

titleElem <- remDr$findElements(using = "class", value = "ratings__user-rating-title")
ratingElem <- remDr$findElements(using = "class", value = "star-display")

#Get the title for all using getElementAttribute
title <- unlist(lapply(titleElem, function(x) {x$getElementAttribute('title')}))

#Start rating will need this helper function to count the stars and half-stars
getStars = function(elem) {
  filledStarsList = elem$findChildElements(using = "class", value = "star-display__filled")
  filledStars <- length(unlist(lapply(filledStarsList, function(x) {x$getElementText()})))
  
  halfStarsList = elem$findChildElements(using = "class", value = "star-display__half")
  halfStars <- unlist(length(unlist(lapply(halfStarsList, function(x) {x$getElementText()}))))
  
  stars = filledStars + (halfStars*.5)
  return(stars)
}

#Apply function of rating elements
rating <- unlist(lapply(ratingElem, getStars))

#combind into 1 data frame
df = cbind(title,rating)

head(df)

#Save this dataframe since the above takes so long, I don't want to do any more than I have to.
write.csv(df,'rotten_tomatoes_ratings.csv',row.names = FALSE)

```


Now I'm going to use the New York Times API to attempt to get the reviews for these movies. This could definitely use some improvement - I'm not cleaning up titles, and i'm just assuming that the first result is the right one.  

In order for this to work, you'll need to set the "apiKey" constant to your api key.

```{r, include=FALSE}
apiKey = 'RvUd6JkD5TqEvKIbpSDqkyTRXUWWGoeU'
```

```{r, eval=FALSE}

#This function will take a row from my dataframe and return a vector with the NYT data appended
getReview = function(row) {
  #Need to sleep for 6 seconds to keep under rate limit
  Sys.sleep(6)
  
  #get title and URL endcode
  titleEncoded = URLencode(row['title'],reserved = TRUE)
  url = paste0('https://api.nytimes.com/svc/movies/v2/reviews/search.json?query=',titleEncoded,'&api-key=',apiKey)
  result <- GET(url)
  warn_for_status(result)
  first_result = NULL
  resultList = fromJSON( content(result,'text'))
  
  #Get the first result, and unlist into a named vector
  if(resultList$num_results > 0) {
    first_result = unlist(resultList$results[1,])
  }  
  #Add the number of results to the vector, so movies with extra results could be examined later.
  first_result['num_results'] = resultList$num_results
  first_result['rt_title']  = row['title']
  first_result['my_rating']  = row['rating']
  
  return(first_result)
    
}

reviewList = apply(df,1,getReview)


#We now have a list of vectors.  Unfortunately they don't all have the same fields, so I'll loop through here to get the list of fields
names = NULL

for (i in 1:length(reviewList)) {
  names = c(names,names(reviewList[[i]]))
}

names = unique(names)
names = names[names!=""]

#Create a new dataframe with our field list
dfnew <- as.data.frame(matrix(,ncol=length(names),nrow=0))
names(dfnew)<-names

#Now I'll loop through our list of vectors to create the final datafame
for(i in 1:length(reviewList)){
  row = reviewList[[i]]
  newRow = NULL
  for(name in names){
    newRow[name] = row[name]
  }
  dfnew = rbind(dfnew,as.data.frame(t(newRow)))
}

head(dfnew)

write.csv(dfnew,'rotten_tomatoes_ratings_with_nyt.csv',row.names = FALSE)

```

```{r, include = FALSE}
dfnew = read.csv('rotten_tomatoes_ratings_with_nyt.csv')
```

OK, so we now that we have a dataframe, let's do some exploration to see how my ratings relate to NYT.

First, let's see if my rating relates to whether or not we found a review at all.

```{r}

dfnew$my_rating <- as.numeric(as.character(dfnew$my_rating))
dfnew$critics_pick<- as.numeric(as.character(dfnew$critics_pick))

grouped_df = group_by(dfnew,my_rating)

rating_summary = summarise(
  grouped_df,
  count = n(),
  missingCount = sum(is.na(display_title)),
  AvgMissing = mean(is.na(display_title))
)


#display sorted summary
rating_summary[order(rating_summary$my_rating),]


ggplot(data=rating_summary, aes(x=my_rating, y=AvgMissing))+
  geom_bar(stat="identity")

```    
  
Hard to say for sure if there is anything of note here, but it does look like the higher I rated a movie, the more likely it is to have a NYT review.  

Next, we can see if my rating correlates to the "critics_pick" flag:  

```{r}
rating_summary_2 = summarise(
  grouped_df,
  count = n(),
  missingCount = sum(is.na(display_title)),
  pickAverage = mean(critics_pick, na.rm = TRUE)
)

#display sorted summary
rating_summary_2[order(rating_summary_2$my_rating),]


ggplot(data=rating_summary_2, aes(x=my_rating, y=pickAverage))+
  geom_bar(stat="identity")

```  

It does appear that there is a strong correlation here. Clearly the NYT critics and I see eye to eye!

